---
title: "Assignment 1: Data Set Selection, Cleaning, and Differential Expression"
author: "Mindy Lee"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
editor_options:
  markdown:
    wrap: 72
---

```{r setup, message=FALSE, warning=FALSE}
# Check and install all required packages before loading
pkgs_cran <- c("data.table", "ggplot2", "pheatmap", "R.utils")
pkgs_bioc <- c("GEOquery", "edgeR", "limma", "HGNChelper")

for (p in pkgs_cran) {
  if (!requireNamespace(p, quietly = TRUE)) install.packages(p)
}
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
for (p in pkgs_bioc) {
  if (!requireNamespace(p, quietly = TRUE)) BiocManager::install(p, ask = FALSE)
}

library(GEOquery)
library(data.table)
library(edgeR)
library(limma)
library(ggplot2)
library(pheatmap)
library(HGNChelper)
library(R.utils)

set.seed(1)
```

# 1. Introduction

Idiopathic Pulmonary Fibrosis (IPF) is a chronic and progressive
interstitial lung disease defined by irreversible scarring of lung
tissue, declining pulmonary function, and poor long-term prognosis.
Macrophages are central regulators of the inflammatory and remodeling
programs that determine whether lung injury resolves or transitions into
pathological fibrosis.

This notebook analyzes **GSE79544**, a human bulk RNA-seq dataset
containing bronchoalveolar lavage (BAL) macrophage-rich samples from
IPF patients and smoking controls. Goals: (i) obtain the dataset
computationally, (ii) assess data quality, (iii) map identifiers to
current HUGO/HGNC symbols, (iv) normalize, and (v) run differential
expression analysis.

The dataset is linked to the publication: Prasse A *et al.* (2018).
*BAL Cell Gene Expression in Idiopathic Pulmonary Fibrosis: A Pilot
Study.* PLOS ONE, 14(1): e0210463. PMID: 29649237.

# 2. Dataset Selection and Justification

GSE79544 was selected because it:

- Uses **human** BAL macrophage-rich samples (meets human-data requirement)
- Compares **IPF patients vs smoking controls** (clear biological contrast)
- Provides **RNA-seq counts** at gene level (transcriptome-wide coverage)
- Has **biological replication** in both groups
- Was submitted to GEO in 2016 (within the ten-year recency window)

# 3. Download Data (Computational, Cached)

```{r download-data}
gse_id   <- "GSE79544"
supp_dir <- file.path("data", gse_id, "supp")
gse_rds  <- file.path("data", gse_id, paste0(gse_id, "_GSEMatrix.rds"))

# Create cache directories
dir.create("data",                        showWarnings = FALSE)
dir.create(file.path("data", gse_id),    showWarnings = FALSE)
dir.create(supp_dir,                      showWarnings = FALSE)

# --- (1) GEO metadata (cached) ---
if (!file.exists(gse_rds)) {
  gse_obj <- GEOquery::getGEO(gse_id, GSEMatrix = TRUE)
  saveRDS(gse_obj, gse_rds)
  message("Downloaded and cached GEO metadata.")
} else {
  gse_obj <- readRDS(gse_rds)
  message("Loaded GEO metadata from cache.")
}

eset  <- gse_obj[[1]]
pheno <- pData(eset)

cat("Metadata samples:", nrow(pheno), "\n")
cat("Metadata columns:", paste(colnames(pheno)[1:6], collapse = ", "), "\n")
```

```{r download-supp}
# --- (2) Supplementary count file (cached) ---
if (length(list.files(supp_dir)) == 0) {
  GEOquery::getGEOSuppFiles(gse_id, baseDir = supp_dir, makeDirectory = FALSE)
  message("Downloaded supplementary files.")
} else {
  message("Supplementary files already cached.")
}

# Identify count file
candidates  <- list.files(supp_dir, full.names = TRUE)
counts_file <- candidates[
  grepl("count", basename(candidates), ignore.case = TRUE) &
  grepl("\\.(txt|tsv|csv)(\\.gz)?$", candidates, ignore.case = TRUE)
]

if (length(counts_file) != 1) {
  message("All supplementary files found:")
  print(basename(candidates))
  stop("Count file not uniquely identified. Adjust pattern.")
}

# Decompress if needed
if (grepl("\\.gz$", counts_file, ignore.case = TRUE)) {
  counts_txt <- sub("\\.gz$", "", counts_file, ignore.case = TRUE)
  if (!file.exists(counts_txt)) {
    R.utils::gunzip(counts_file, destname = counts_txt, overwrite = TRUE)
  }
} else {
  counts_txt <- counts_file
}

cat("Using count file:", basename(counts_txt), "\n")
```

```{r load-counts}
counts_dt <- data.table::fread(counts_txt)
cat("Raw counts dimensions:", paste(dim(counts_dt), collapse = " x "), "\n")

knitr::kable(
  head(counts_dt[, 1:5]),
  caption = "Table 3.1: Preview of raw count table (first 6 genes, first 4 samples)."
)
```

**Table 3.1 legend:** Small subset of the raw RNA-seq count matrix as
downloaded from GEO supplementary files. Rows = genes; columns =
individual RNA-seq libraries. Values are raw read counts prior to
filtering and normalization.

# 4. Construct Count Matrix

```{r build-matrix}
gene_col   <- names(counts_dt)[1]
counts_df  <- as.data.frame(counts_dt)

rownames(counts_df)  <- counts_df[[gene_col]]
counts_df[[gene_col]] <- NULL

counts_mat <- as.matrix(counts_df)
storage.mode(counts_mat) <- "numeric"

cat("Count matrix dims:", paste(dim(counts_mat), collapse = " x "), "\n")
cat("Sample names in count matrix:\n")
print(colnames(counts_mat))
```

# 5. Define Conditions (Control vs IPF)

```{r define-conditions}
# Print sample titles to see the raw metadata
cat("Sample titles from GEO metadata:\n")
print(pheno$title)
```

```{r assign-conditions}
# Assign condition using the sample title field.
# We check for "Control" or "IPF" in the title (case-insensitive).
# Samples matching neither label are excluded.

pheno$condition <- ifelse(
  grepl("Control", pheno$title, ignore.case = TRUE), "Control",
  ifelse(
    grepl("IPF", pheno$title, ignore.case = TRUE), "IPF",
    NA
  )
)

cat("Condition assignment from title matching:\n")
print(table(pheno$condition, useNA = "always"))
```

```{r align-counts-pheno}
# ------------------------------------------------------------------
# The count matrix columns look like:  CS1_CTR, CS3_IPF, etc.
# The pheno short names look like:     CS1, CS3, etc.
# Strategy: strip the _CTR / _IPF suffix from count column names
# to get a bare ID, then match that to pheno short names.
# ------------------------------------------------------------------

pheno_clean <- pheno[!is.na(pheno$condition), ]
pheno_clean$condition <- factor(pheno_clean$condition,
                                levels = c("Control", "IPF"))

# Build short name from the first whitespace-delimited token of title
pheno_clean$short_name <- gsub("^(\\S+).*", "\\1", pheno_clean$title)

cat("Pheno short names:\n")
print(pheno_clean$short_name)

cat("\nCount matrix column names:\n")
print(colnames(counts_mat))

# Strip _CTR or _IPF suffix from count column names to get bare IDs
count_bare <- gsub("_(CTR|IPF)$", "", colnames(counts_mat),
                   ignore.case = TRUE)

cat("\nBare count IDs after stripping suffix:\n")
print(count_bare)

# Build a named vector: bare_id -> count_column_name
bare_to_col <- setNames(colnames(counts_mat), count_bare)

# Match bare count IDs to pheno short names
matched_cols <- bare_to_col[pheno_clean$short_name]

cat("\nMatched count columns per pheno row:\n")
print(matched_cols)

# Check how many matched
n_matched <- sum(!is.na(matched_cols))
cat("\nMatched", n_matched, "of", nrow(pheno_clean), "pheno rows\n")

if (n_matched == 0) {
  stop(paste(
    "Still no matches after suffix stripping.",
    "Bare count IDs:", paste(count_bare, collapse = ", "),
    "Pheno short names:", paste(pheno_clean$short_name, collapse = ", ")
  ))
}

# Keep only rows that matched
pheno_clean   <- pheno_clean[!is.na(matched_cols), ]
matched_cols  <- matched_cols[!is.na(matched_cols)]

# Subset and reorder count matrix
counts_mat  <- counts_mat[, matched_cols, drop = FALSE]

# Rename count matrix columns to the bare IDs so downstream
# code sees consistent names
colnames(counts_mat) <- names(matched_cols)

# Final safety check
stopifnot(ncol(counts_mat) == nrow(pheno_clean))
cat("Samples after alignment:", ncol(counts_mat), "\n")

knitr::kable(
  data.frame(
    count_col = colnames(counts_mat),
    condition = as.character(pheno_clean$condition)
  ),
  caption = "Table 5.1: Verified sample-to-condition assignment."
)

```
**Table 5.1 legend:** Number of biological samples per condition after
parsing GEO metadata and aligning to the count matrix. These group
labels define the comparison tested in differential expression analysis.

```{r verify-alignment}
# Print a verification table so the reader can confirm assignments
verify_df <- data.frame(
  count_col = colnames(counts_mat),
  condition = as.character(pheno_clean$condition),
  stringsAsFactors = FALSE
)

knitr::kable(
  verify_df,
  caption = "Table 5.2: Verified sample-to-condition assignment."
)
```

**Table 5.2 legend:** Each row confirms which condition label has been
assigned to each count matrix column. Inspect this table to catch any
misassignment before proceeding.

# 6. Exploratory Assessment (QC Before Normalization)

## 6.1 Library Size Distribution

```{r library-sizes}
lib_sizes <- colSums(counts_mat)

hist(
  lib_sizes,
  breaks = 20,
  main   = "Figure 6.1: Distribution of raw library sizes",
  xlab   = "Total reads per sample",
  col    = "steelblue",
  border = "white"
)
```

**Figure 6.1 legend:** Histogram of total mapped reads per sample
before normalization. Variation in library size motivates TMM
normalization so that expression comparisons are not confounded by
sequencing depth differences.

```{r lib-size-table}
knitr::kable(
  data.frame(
    Sample    = colnames(counts_mat),
    Condition = as.character(pheno_clean$condition),
    LibSize   = lib_sizes
  ),
  caption = "Table 6.1: Raw library sizes per sample."
)
```

## 6.2 Density Plot of log2 Raw Counts

```{r density-raw}
log_raw <- log2(counts_mat + 1)

# Colour by condition
cond_colours <- ifelse(pheno_clean$condition == "Control",
                       "steelblue", "firebrick")

plot(
  density(log_raw[, 1]),
  main  = "Figure 6.2: Density of log2(raw counts + 1) per sample",
  xlab  = "log2(count + 1)",
  col   = cond_colours[1],
  ylim  = c(0, 0.25)
)
for (i in 2:ncol(log_raw)) {
  lines(density(log_raw[, i]), col = cond_colours[i])
}
legend("topright",
       legend = c("Control", "IPF"),
       col    = c("steelblue", "firebrick"),
       lty    = 1, bty = "n")
```

**Figure 6.2 legend:** Density curves of log2-transformed raw counts
across all samples. Colour indicates condition (blue = Control, red =
IPF). Misalignment between curves indicates library composition
differences that normalization should correct.

# 7. Map to Current HUGO/HGNC Symbols

```{r hgnc-mapping}
# Map gene identifiers to current HUGO symbols using HGNChelper.
# Rules:
#   (a) If HGNChelper suggests a current symbol, use it.
#   (b) If not, keep the original label.
#   (c) Rows with empty/NA symbols after mapping are removed.
#   (d) Duplicate symbols are collapsed by summing counts.

mapped <- HGNChelper::checkGeneSymbols(rownames(counts_mat))

new_sym <- ifelse(
  is.na(mapped$Suggested.Symbol) | mapped$Suggested.Symbol == "",
  mapped$x,
  mapped$Suggested.Symbol
)

rownames(counts_mat) <- new_sym

# Remove unmapped / empty rows
n_before_unmapped <- nrow(counts_mat)
counts_mat <- counts_mat[
  !is.na(rownames(counts_mat)) & rownames(counts_mat) != "",
  , drop = FALSE
]
n_after_unmapped <- nrow(counts_mat)

# Collapse duplicates by summing
n_before_dup <- nrow(counts_mat)
counts_mat   <- rowsum(counts_mat, group = rownames(counts_mat))
n_after_dup  <- nrow(counts_mat)

cat("Rows removed (unmapped/empty)  :", n_before_unmapped - n_after_unmapped, "\n")
cat("Rows collapsed (duplicates)    :", n_before_dup      - n_after_dup,      "\n")
cat("Genes after HGNC cleanup       :", nrow(counts_mat),                     "\n")
```

**Mapping interpretation:** Multiple rows mapping to the same HUGO
symbol were collapsed by summation because raw counts are additive
across transcript fragments targeting the same gene. Rows with no
valid symbol were dropped to prevent ambiguous identifiers from
propagating into downstream steps.

# 8. Filtering and Normalization

## 8.1 Why Filtering?

Genes with very low counts across all samples contribute negligible
statistical information while inflating the multiple-testing burden.
`filterByExpr` retains genes that achieve a minimum count-per-million
(CPM) in a number of samples at least equal to the smallest group size.

## 8.2 Why TMM Normalization?

TMM (Trimmed Mean of M-values) corrects for library composition bias,
where a few highly expressed genes can distort per-sample scaling
factors. It is the recommended normalization for edgeR-based RNA-seq
workflows (Robinson & Oshlack, 2010).

```{r filtering-normalization}
# Build DGEList
dge <- DGEList(counts = counts_mat,
               group  = pheno_clean$condition)

# Filter lowly expressed genes
keep <- filterByExpr(dge, group = pheno_clean$condition)
dge  <- dge[keep, , keep.lib.sizes = FALSE]
cat("Genes retained after filtering:", nrow(dge), "\n")

# TMM normalization
dge    <- calcNormFactors(dge, method = "TMM")
logcpm <- cpm(dge, log = TRUE, prior.count = 2)

knitr::kable(
  data.frame(
    Sample     = rownames(dge$samples),
    Condition  = as.character(dge$samples$group),
    LibSize    = dge$samples$lib.size,
    NormFactor = round(dge$samples$norm.factors, 4)
  ),
  caption = "Table 8.1: Library sizes and TMM normalization factors."
)
```

**Table 8.1 legend:** TMM normalization factors close to 1.0 indicate
little composition bias. Values substantially above or below 1.0 flag
samples where a small number of dominant transcripts would otherwise
distort fold-change estimates.

## 8.3 Pre vs Post Normalization Boxplots

```{r norm-boxplots, fig.width=12, fig.height=5}
par(mfrow = c(1, 2), mar = c(8, 4, 4, 2))

boxplot(
  log2(dge$counts + 1),
  main    = "Figure 8.1: Pre-normalization\nlog2(counts + 1)",
  outline = FALSE,
  las     = 2,
  col     = ifelse(pheno_clean$condition == "Control",
                   "steelblue", "firebrick"),
  cex.axis = 0.6
)

boxplot(
  logcpm,
  main    = "Figure 8.2: Post-normalization\nlogCPM (TMM)",
  outline = FALSE,
  las     = 2,
  col     = ifelse(pheno_clean$condition == "Control",
                   "steelblue", "firebrick"),
  cex.axis = 0.6
)

par(mfrow = c(1, 1))
```

**Figures 8.1–8.2 legend:** Boxplots summarize per-sample expression
distributions before (left) and after (right) TMM normalization. After
normalization the median logCPM values are aligned across samples,
indicating that library-size differences have been corrected. Blue =
Control, red = IPF.

```{r density-post-norm}
cond_colours <- ifelse(pheno_clean$condition == "Control",
                       "steelblue", "firebrick")

plot(
  density(logcpm[, 1]),
  main  = "Figure 8.3: Density of logCPM after TMM normalization",
  xlab  = "logCPM",
  col   = cond_colours[1],
  ylim  = c(0, 0.25)
)
for (i in 2:ncol(logcpm)) {
  lines(density(logcpm[, i]), col = cond_colours[i])
}
legend("topright",
       legend = c("Control", "IPF"),
       col    = c("steelblue", "firebrick"),
       lty    = 1, bty = "n")
```

**Figure 8.3 legend:** Density curves of normalized logCPM values.
Improved alignment relative to Figure 6.2 confirms that TMM has
reduced library-driven differences while retaining biological
variability.

# 9. Structure and Outlier Assessment (MDS + PCA)

```{r mds-plot}
cond_colours <- ifelse(pheno_clean$condition == "Control",
                       "steelblue", "firebrick")

plotMDS(
  dge,
  col    = cond_colours,
  pch    = 19,
  main   = "Figure 9.1: MDS plot (leading log-fold-change distances)"
)
legend("topright",
       legend = c("Control", "IPF"),
       col    = c("steelblue", "firebrick"),
       pch    = 19, bty = "n")
```

**Figure 9.1 legend:** MDS plot summarises sample-to-sample distances
based on the top genes with largest expression differences. Separation
between IPF and Control along dimension 1 indicates condition-driven
transcriptomic structure. Any sample falling far from its group would
be investigated as a potential outlier.

```{r pca-plot}
pca <- prcomp(t(logcpm), scale. = TRUE)

pct_var <- round(100 * pca$sdev^2 / sum(pca$sdev^2), 1)

plot(
  pca$x[, 1], pca$x[, 2],
  col  = cond_colours,
  pch  = 19,
  xlab = paste0("PC1 (", pct_var[1], "% variance)"),
  ylab = paste0("PC2 (", pct_var[2], "% variance)"),
  main = "Figure 9.2: PCA of TMM-normalized expression"
)
text(pca$x[, 1], pca$x[, 2],
     labels = colnames(logcpm),
     cex    = 0.6,
     pos    = 3)
legend("topright",
       legend = c("Control", "IPF"),
       col    = c("steelblue", "firebrick"),
       pch    = 19, bty = "n")
```

**Figure 9.2 legend:** PCA displays the major axes of transcriptomic
variability. Separation of IPF from Control samples along early
principal components supports that disease status explains a meaningful
fraction of expression variance. Sample labels allow identification of
any individual outliers.

```{r outlier-decision}
cat("Outlier decision:\n")
cat("No samples were removed. Both MDS and PCA show reasonable\n")
cat("group clustering without extreme outliers. Biological\n")
cat("heterogeneity is expected in human patient cohorts.\n")
```

# 10. Differential Expression Analysis (limma-voom)

limma-voom is used because it handles the mean-variance relationship
inherent in RNA-seq count data by computing observation-level precision
weights, enabling reliable linear modelling with moderate sample sizes.

```{r voom-design}
# Design matrix: no intercept, one coefficient per condition
design <- model.matrix(~ 0 + pheno_clean$condition)
colnames(design) <- levels(pheno_clean$condition)

cat("Design matrix (first 6 rows):\n")
print(head(design))
```

```{r voom-fit}
# voom transformation with mean-variance plot
v <- voom(dge, design, plot = TRUE)
title("Figure 10.1: voom mean-variance trend")
```

**Figure 10.1 legend:** The voom mean-variance plot shows the
empirical trend used to compute observation-level weights. A smooth,
monotonically decreasing trend (higher variance at low counts)
confirms the model is appropriate for these data.

```{r lmfit}
fit  <- lmFit(v, design)
cont <- makeContrasts(IPFvsControl = IPF - Control, levels = design)
fit2 <- eBayes(contrasts.fit(fit, cont))

# Extract full results table
tt       <- topTable(fit2, number = Inf, sort.by = "P")
tt$gene  <- rownames(tt)

cat("Total genes tested:", nrow(tt), "\n")
```

# 11. Significance Thresholds and Multiple Testing

Two criteria are applied to define significant differential expression:

1. **FDR < 0.05** using Benjamini-Hochberg correction — necessary
   because thousands of genes are tested simultaneously.
2. **|log2FC| >= 1** — requires at least a ~2-fold change to focus on
   biologically meaningful effect sizes.

```{r de-thresholds}
tt$signif <- tt$adj.P.Val < 0.05 & abs(tt$logFC) >= 1

n_up   <- sum(tt$signif & tt$logFC  > 0)
n_down <- sum(tt$signif & tt$logFC  < 0)
n_total <- sum(tt$signif)

cat(sprintf("Up-regulated   (FDR<0.05, logFC> 1): %d\n", n_up))
cat(sprintf("Down-regulated (FDR<0.05, logFC<-1): %d\n", n_down))
cat(sprintf("Total significant DEGs             : %d\n", n_total))
```

```{r de-top-table}
knitr::kable(
  head(tt[, c("gene","logFC","AveExpr","t","P.Value","adj.P.Val","B")], 15),
  digits  = 4,
  caption = "Table 11.1: Top 15 differentially expressed genes (sorted by p-value)."
)
```

**Table 11.1 legend:** The 15 genes with the smallest unadjusted
p-values from the IPF vs Control limma-voom analysis. logFC > 0 =
higher in IPF; logFC < 0 = lower in IPF. adj.P.Val is the
Benjamini-Hochberg FDR.

# 12. Visualizations

## 12.1 Volcano Plot

```{r volcano-plot, fig.width=8, fig.height=6}
# Classify genes for colouring
tt$direction <- ifelse(
  tt$adj.P.Val < 0.05 & tt$logFC >  1, "Up",
  ifelse(
    tt$adj.P.Val < 0.05 & tt$logFC < -1, "Down",
    "NS"
  )
)

# Label the top 20 most significant genes
label_df <- tt[order(tt$adj.P.Val), ][1:20, ]

ggplot(tt, aes(x = logFC, y = -log10(adj.P.Val), colour = direction)) +
  geom_point(alpha = 0.5, size = 1.2) +
  scale_colour_manual(
    values = c(Up = "#D7191C", Down = "#2C7BB6", NS = "grey70"),
    labels = c(Up = "Up in IPF", Down = "Down in IPF", NS = "Not significant")
  ) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", alpha = 0.6) +
  geom_vline(xintercept = c(-1, 1),     linetype = "dashed", alpha = 0.6) +
  geom_text(
    data  = label_df,
    aes(label = gene),
    size  = 2.5, vjust = -0.5, colour = "black"
  ) +
  labs(
    title    = "Figure 12.1: Volcano plot – IPF vs Control",
    subtitle = sprintf("%d up, %d down (FDR < 0.05, |logFC| >= 1)",
                       n_up, n_down),
    x        = "log2 Fold Change (IPF / Control)",
    y        = "-log10(FDR)",
    colour   = NULL
  ) +
  theme_bw(base_size = 12)
```

**Figure 12.1 legend:** Volcano plot combining effect size (log2FC)
and statistical significance (FDR). Red points are genes significantly
up-regulated in IPF; blue are down-regulated; grey are not significant.
Dashed lines mark |logFC| = 1 and FDR = 0.05. The 20 most significant
genes are labelled.

## 12.2 MA Plot

```{r ma-plot}
plotMA(fit2,
       main   = "Figure 12.2: MA plot – IPF vs Control",
       status = tt$direction[match(rownames(fit2), tt$gene)],
       values = c("Up", "Down", "NS"),
       col    = c("#D7191C", "#2C7BB6", "grey70"))
```

**Figure 12.2 legend:** MA plot shows log fold-change (M) against
average expression (A). The horizontal line at M = 0 represents no
change. Most genes cluster around zero fold-change; highlighted points
pass the significance threshold.

## 12.3 Heatmap of Top Genes

```{r heatmap, fig.width=9, fig.height=8}
# Use top 50 genes by p-value regardless of significance threshold
# so the heatmap is always populated even with strict FDR
top50 <- tt$gene[order(tt$P.Value)][1:min(50, nrow(tt))]
mat50 <- logcpm[top50, , drop = FALSE]

annotation_col    <- data.frame(Condition = pheno_clean$condition)
rownames(annotation_col) <- colnames(mat50)

annotation_colors <- list(
  Condition = c(Control = "#4DAF4A", IPF = "#E41A1C")
)

pheatmap::pheatmap(
  mat50,
  scale             = "row",
  annotation_col    = annotation_col,
  annotation_colors = annotation_colors,
  show_colnames     = FALSE,
  fontsize_row      = 7,
  main              = "Figure 12.3: Top 50 genes by p-value (row-scaled logCPM)"
)
```

**Figure 12.3 legend:** Heatmap of the 50 genes with the smallest
p-values. Row-scaling (z-score) emphasises relative expression
patterns. Hierarchical clustering of columns reveals whether IPF and
Control samples group by condition. Colour bars at the top indicate
condition assignment (green = Control, red = IPF).

# 13. Final Dataset Object

The assignment requires a normalized expression matrix with unique
HUGO symbols as rownames and numeric expression values as columns.

```{r final-dataset}
final_dataset <- logcpm

# Verify requirements
stopifnot(
  length(unique(rownames(final_dataset))) == nrow(final_dataset),
  is.numeric(final_dataset[1, 1])
)

cat("Final dataset dimensions (genes x samples):",
    paste(dim(final_dataset), collapse = " x "), "\n")
cat("Unique rownames (HUGO symbols):", length(unique(rownames(final_dataset))), "\n")

knitr::kable(
  head(final_dataset[, 1:5]),
  caption = "Table 13.1: Preview of final normalized dataset (logCPM, TMM-normalized)."
)
```

**Table 13.1 legend:** First six rows and five columns of the final
normalized expression matrix. Each entry is a log2-CPM value after TMM
normalization. Rownames are unique current HUGO gene symbols.

# 14. Save DE Results for Assignment 2

```{r save-de-for-a2}
# Save the DE results table in the exact column format expected by
# A2_MindyLee.Rmd:
#   SYMBOL  - HUGO gene symbol
#   logFC   - log2 fold change (IPF vs Control)
#   logCPM  - average log2 CPM (AveExpr from topTable)
#   t       - moderated t-statistic
#   PValue  - raw p-value
#   FDR     - Benjamini-Hochberg adjusted p-value

de_for_a2 <- data.frame(
  SYMBOL = tt$gene,
  logFC  = tt$logFC,
  logCPM = tt$AveExpr,
  t      = tt$t,
  PValue = tt$P.Value,
  FDR    = tt$adj.P.Val,
  stringsAsFactors = FALSE
)

saveRDS(de_for_a2,
        file = file.path("data", "GSE79544", "de_results.rds"))

cat("Saved de_results.rds\n")
cat("Rows :", nrow(de_for_a2), "\n")
cat("Columns:", paste(colnames(de_for_a2), collapse = ", "), "\n")
cat("\nDE summary:\n")
cat("  Up-regulated   (FDR<0.05, logFC> 1):", n_up,    "\n")
cat("  Down-regulated (FDR<0.05, logFC<-1):", n_down,  "\n")
cat("  Total significant                   :", n_total, "\n")
```

# 15. Explicit Answers to Required Questions

**Why is the dataset of interest?**
GSE79544 profiles macrophage-rich BAL samples from IPF patients and
smoking controls. Macrophages are central to fibrotic signalling, making
their transcriptomes a biologically meaningful readout of disease-
associated immune programs.

**What are the control and test conditions?**
Control = smoking controls; Test = IPF patients. Both groups provided
BAL macrophage-enriched samples.

**How many samples per condition?**
See Table 5.1. Control n = 7, IPF n = 16.

**Were there expression values not unique for specific genes?**
Yes. Multiple rows mapped to the same HUGO symbol after HGNChelper
cleanup. These were collapsed by summing raw counts (Section 7).

**Were there expression values that could not be mapped to HUGO symbols?**
Rows with empty or NA symbols after HGNChelper mapping were removed
(count reported in Section 7 output).

**Were there any outliers? How many were removed?**
No samples were removed. MDS and PCA (Figures 9.1–9.2) showed
reasonable group clustering without extreme outliers. The originating
paper (Prasse *et al.*, 2018) did not report excluding any of the 23
samples present in this GEO dataset.

**How were replicates handled?**
Biological replicates within each condition are retained as separate
columns. They contribute to within-group variance estimation in the
limma-voom linear model.

**What is the final coverage?**
`r nrow(final_dataset)` genes after HGNC mapping, duplicate collapsing,
low-expression filtering, and TMM normalization.

# 16. Conclusion

This notebook implements a complete, reproducible workflow producing a
clean, TMM-normalized expression matrix with unique HUGO gene symbols
for GSE79544. Key steps: computational download and caching, library-
size QC, HGNC symbol mapping with explicit duplicate handling, TMM
normalization, MDS/PCA-based outlier assessment, and limma-voom
differential expression. The final dataset (`final_dataset`) and DE
results (`de_results.rds`) are ready for Assignment 2 pathway and
network analysis.

# 17. References

- Prasse A *et al.* (2018). BAL Cell Gene Expression in IPF. *PLOS ONE* 14(1):e0210463.
- Ritchie ME *et al.* (2015). limma. *Nucleic Acids Res* 43(7):e47.
- Robinson MD, McCarthy DJ, Smyth GK (2010). edgeR. *Bioinformatics* 26:139–140.
- Robinson MD, Oshlack A (2010). TMM normalization. *Genome Biology* 11:R25.
- Davis S, Meltzer PS (2007). GEOquery. *Bioinformatics* 23:1846–1847.
- Benjamini Y, Hochberg Y (1995). Controlling the FDR. *JRSS-B* 57:289–300.
- Kolde R (2019). pheatmap R package. https://CRAN.R-project.org/package=pheatmap
```