---
title: "Assignment 1: Data Set Selection, Cleaning, Normalization, and Differential Expression"
author: "Mindy Lee"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)
set.seed(1)
```

# Introduction

Idiopathic Pulmonary Fibrosis (IPF) is a chronic and progressive interstitial lung disease defined by irreversible scarring of lung tissue, declining pulmonary function, and poor long-term prognosis. While fibrosis is often described as a problem of extracellular matrix deposition, it is increasingly understood as a systems-level failure of injury response: inflammatory signaling persists, repair programs do not resolve properly, and tissue remodeling becomes self-reinforcing. In this context, immune cells are not merely bystanders—they can help determine whether injury resolves or transitions into pathological fibrosis.

Macrophages are especially relevant to IPF biology because they integrate signals from damaged epithelium, pathogens or irritants, and stromal remodeling. In the lung, macrophage states can shift toward pro-inflammatory or pro-fibrotic phenotypes, influencing cytokine production, growth factor signaling, and extracellular matrix regulation. Therefore, profiling macrophage-rich bronchoalveolar lavage (BAL) samples provides a biologically grounded way to study immune programs that correlate with disease status in human subjects.

In this notebook, I analyze GEO dataset **GSE79544**, a human bulk RNA-seq dataset containing BAL macrophage-rich samples from IPF patients and smoking controls. The goals of this assignment are to (i) obtain the dataset computationally, (ii) assess data quality and overall structure, (iii) map gene identifiers to **unique current HUGO/HGNC symbols**, (iv) clean and normalize the dataset into a reusable expression matrix, and (v) conduct differential expression analysis using a model justified by exploratory plots (MDS/PCA). The final product is a normalized expression matrix with unique HUGO symbols as rownames and numeric expression values in columns, suitable for subsequent assignments (scoring, enrichment, etc.).

# Dataset selection and justification

The dataset **GSE79544** was selected because it satisfies the assignment requirements while also representing an experiment with clear biological meaning. First, the data are derived from **human BAL macrophage-rich samples**, meeting the requirement for human tissue/cell expression data. Second, the experimental comparison is interpretable: IPF patients are contrasted against smoking controls, enabling differential expression in the context of a clinically relevant lung exposure background. Third, the dataset provides RNA-seq counts at gene level, which supports transcriptome-wide coverage rather than targeted measurement. Finally, the sample size (Control n≈7, IPF n≈16 as parsed from metadata) provides biological replication in both groups, making dispersion estimation and downstream modeling feasible.

From an analytical standpoint, this dataset is appropriate for a complete, reproducible workflow: raw counts are available through GEO supplementary files, sample annotations can be obtained through GEOquery, and identifiers can be mapped to current HGNC symbols with a transparent set of rules for handling unmapped and duplicated genes.

GSE79544 also meets the course “recency/quality” requirement: the GEO series was submitted on **March 23, 2016**, placing it within the last ten years. The dataset is linked to a primary publication describing experimental context and processing, “Bronchoalveolar lavage (BAL) cells in idiopathic pulmonary fibrosis express a complex pro-inflammatory, pro-repair, angiogenic activation pattern…” (PLOS ONE, 2018; PMID: 29649237), which I cite throughout this report for provenance.


# Install packages

```{r packages, results='hide', message=FALSE, warning=FALSE}
pkgs_cran <- c("data.table", "ggplot2", "pheatmap", "R.utils")
pkgs_bioc <- c("GEOquery", "edgeR", "limma", "HGNChelper")

for (p in pkgs_cran) {
  if (!requireNamespace(p, quietly = TRUE)) install.packages(p)
}
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
for (p in pkgs_bioc) {
  if (!requireNamespace(p, quietly = TRUE)) BiocManager::install(p, ask = FALSE)
}

library(GEOquery)
library(data.table)
library(edgeR)
library(limma)
library(ggplot2)
library(pheatmap)
library(HGNChelper)
library(R.utils)
```

# Download data (computational, cached)

This section downloads the GEO Series Matrix (metadata) and the supplementary files containing the count matrix. To avoid re-downloading during repeated knitting, the notebook caches both the metadata object and the extracted count file in the `data/` directory.

```{r download_data}
gse_id <- "GSE79544"

dir.create("data", showWarnings = FALSE)
dir.create(file.path("data", gse_id), showWarnings = FALSE)
dir.create(file.path("data", gse_id, "supp"), showWarnings = FALSE)

gse_rds  <- file.path("data", gse_id, paste0(gse_id, "_GSEMatrix.rds"))
supp_dir <- file.path("data", gse_id, "supp")

# (1) GEO Series Matrix / metadata (cached)
if (!file.exists(gse_rds)) {
  gse_obj <- GEOquery::getGEO(gse_id, GSEMatrix = TRUE)
  saveRDS(gse_obj, gse_rds)
} else {
  gse_obj <- readRDS(gse_rds)
}

eset  <- gse_obj[[1]]
pheno <- pData(eset)

cat("Metadata samples:", nrow(pheno), "\n")

# (2) Supplementary files (cached)
if (length(list.files(supp_dir)) == 0) {
  GEOquery::getGEOSuppFiles(gse_id, baseDir = supp_dir, makeDirectory = FALSE)
}

# Identify a likely counts file.
# If this fails on your machine, print list.files(supp_dir) and adjust patterns.
candidates <- list.files(supp_dir, full.names = TRUE)

# Prefer .txt/.tsv + optional .gz, containing "count" or "counts"
counts_file <- candidates[grepl("count", basename(candidates), ignore.case = TRUE) &
                            grepl("\\.(txt|tsv|csv)(\\.gz)?$", candidates, ignore.case = TRUE)]

if (length(counts_file) != 1) {
  message("Found these potential count files:")
  print(basename(counts_file))
  message("All supplementary files:")
  print(basename(candidates))
  stop("Counts file not uniquely identified. Adjust the pattern above to match the correct file name.")
}

# Unzip if needed
if (grepl("\\.gz$", counts_file, ignore.case = TRUE)) {
  counts_txt <- sub("\\.gz$", "", counts_file, ignore.case = TRUE)
  if (!file.exists(counts_txt)) {
    R.utils::gunzip(counts_file, destname = counts_txt, overwrite = TRUE)
  }
} else {
  counts_txt <- counts_file
}

cat("Using count file:", basename(counts_txt), "\n")

counts_dt <- data.table::fread(counts_txt)
cat("Counts table dims:", paste(dim(counts_dt), collapse = " x "), "\n")

knitr::kable(
  head(counts_dt[, 1:5]),
  caption = "Table 4.1: Preview of the raw count table (first 6 genes, first 4 samples)."
)
```

**Table 4.1 legend:**  
This table shows a small subset of the raw RNA-seq count matrix as downloaded from GEO supplementary files. Rows correspond to genes (as labeled in the file) and columns correspond to individual RNA-seq libraries (samples). The numeric values are raw read counts prior to filtering and normalization. Because sequencing depth and library composition can vary across samples, these raw counts are not directly comparable between samples without normalization.

# Construct count matrix (required for downstream steps)

The GEO supplementary counts are loaded into a matrix format suitable for RNA-seq modeling. The first column is treated as the gene identifier, and all remaining columns are interpreted as sample count columns.

```{r build_counts_mat}
gene_col <- names(counts_dt)[1]
counts_df <- as.data.frame(counts_dt)

rownames(counts_df) <- counts_df[[gene_col]]
counts_df[[gene_col]] <- NULL

counts_mat <- as.matrix(counts_df)
storage.mode(counts_mat) <- "numeric"

cat("Counts matrix dims:", paste(dim(counts_mat), collapse = " x "), "\n")
```

**Interpretation of the raw count preview.**  
Table X shows a small slice of the downloaded count matrix. Each row corresponds to a gene identifier provided by the dataset authors, and each column is one RNA-seq library (one patient sample). The values are **raw mapped read counts**, meaning they reflect both (i) true biology (genes genuinely more/less expressed) and (ii) technical factors such as sequencing depth and library composition. Because different samples can have different total read counts, raw values are not directly comparable across columns; therefore, downstream steps apply filtering and normalization so that group comparisons (IPF vs control) are not driven by depth artifacts.


# Define conditions (Control vs IPF)

The phenotype metadata is used to define the primary biological factor for analysis: **condition** (Control vs IPF). The rule below is based on pattern matching in the GEO sample titles; this is a transparent and reproducible approach, but it should always be sanity-checked against the GEO record.

```{r define_groups}
pheno$condition <- ifelse(
  grepl("IPF", pheno$title, ignore.case = TRUE), "IPF",
  ifelse(grepl("Control", pheno$title, ignore.case = TRUE), "Control", NA)
)

pheno_clean <- pheno[!is.na(pheno$condition), ]
pheno_clean$condition <- factor(pheno_clean$condition, levels = c("Control", "IPF"))

knitr::kable(
  as.data.frame(table(pheno_clean$condition)),
  col.names = c("Group", "Count"),
  caption = "Table 6.1: Sample counts per condition (Control vs IPF)."
)
```

**Table 6.1 interpretation:**  
This table summarizes the number of biological samples used in each condition after parsing metadata. These group labels define the comparison tested in differential expression analysis. If a sample cannot be confidently assigned, it is excluded rather than guessed.

# Exploratory assessment (QC before normalization)

This section assesses technical variability and overall structure prior to normalization. The goal is not to “judge significance,” but to determine whether samples have consistent depth and comparable distributions.

## Library size distribution

```{r qc_library_sizes}
lib_sizes <- colSums(counts_mat)

hist(lib_sizes,
     breaks = 20,
     main = "Figure 7.1: Distribution of raw library sizes",
     xlab = "Total reads per sample")

knitr::kable(
  data.frame(
    sample = colnames(counts_mat)[1:min(10, ncol(counts_mat))],
    lib_size = lib_sizes[1:min(10, length(lib_sizes))]
  ),
  caption = "Table 7.1: Raw library sizes (first 10 samples)."
)
```

**Figure 7.1 legend:**  
Histogram of raw library sizes (total reads per sample). Variation is expected in RNA-seq, but substantial differences motivate normalization to ensure that observed expression differences reflect biology rather than sequencing depth.

## Density plot of log2 raw counts

```{r qc_raw_density}
log_raw <- log2(counts_mat + 1)

plot(density(log_raw[, 1]),
     main = "Figure 7.2: Density of log2(raw counts + 1)",
     xlab = "log2(count + 1)")

for (i in 2:ncol(log_raw)) {
  lines(density(log_raw[, i]), col = i)
}
```

**Figure 7.2 legend:**  
Density curves of log2-transformed raw counts across samples. Misalignment between curves suggests differences in library composition and depth. Normalization aims to reduce these technical differences while preserving true biological signal.

# Map to current HUGO/HGNC symbols (Map + Clean)

RNA-seq count matrices can include outdated gene symbols, aliases, or identifiers that are not current HUGO symbols. To ensure that the final dataset is compatible with downstream annotation and pathway analysis, gene identifiers are cleaned using **HGNChelper**. The rules below are explicit: if HGNChelper suggests a current symbol, it is used; otherwise the original label is retained. Unmappable/empty symbols are removed. Duplicate symbols are collapsed by summing counts, producing one row per unique symbol.

```{r hgnc_cleanup}
mapped <- HGNChelper::checkGeneSymbols(rownames(counts_mat))

new_sym <- ifelse(
  is.na(mapped$Suggested.Symbol) | mapped$Suggested.Symbol == "",
  mapped$x,
  mapped$Suggested.Symbol
)

rownames(counts_mat) <- new_sym

before_unmapped <- nrow(counts_mat)
counts_mat <- counts_mat[rownames(counts_mat) != "" & !is.na(rownames(counts_mat)), , drop = FALSE]
after_unmapped <- nrow(counts_mat)

dup_before <- nrow(counts_mat)
counts_mat <- rowsum(counts_mat, group = rownames(counts_mat))
dup_after <- nrow(counts_mat)

cat("Rows removed due to unmapped/empty symbols:", before_unmapped - after_unmapped, "\n")
cat("Rows collapsed due to duplicates:", dup_before - dup_after, "\n")
cat("Post-HGNC cleanup dims:", paste(dim(counts_mat), collapse = " x "), "\n")
```

**Interpretation (mapping + duplicates):**  
Rows that could not be associated with a valid gene symbol (empty/NA after cleanup) were removed to prevent ambiguous identifiers from propagating into downstream steps. Multiple rows mapping to the same symbol were collapsed by summation because counts are additive across fragments/probes targeting the same gene. This results in a final expression set with unique gene symbols as rownames, satisfying the assignment specification.

# Filtering and normalization

## Why filtering?

Very lowly expressed genes contribute little statistical power while inflating the multiple-testing burden. Filtering reduces noise and improves the reliability of variance estimation.

## Why TMM normalization?

TMM (Trimmed Mean of M-values) is widely used for RNA-seq because it corrects for library composition bias, where a subset of highly expressed genes can distort scaling. TMM is robust in differential expression settings where genuine biological shifts exist (Robinson & Oshlack, 2010).

```{r normalization}
# Ensure column names exist and match samples
if (is.null(colnames(counts_mat)) || any(colnames(counts_mat) == "")) {
  colnames(counts_mat) <- pheno_clean$geo_accession
}
stopifnot(ncol(counts_mat) == nrow(pheno_clean))

dge <- DGEList(counts = counts_mat, group = pheno_clean$condition)

keep <- filterByExpr(dge, group = pheno_clean$condition)
dge <- dge[keep, , keep.lib.sizes = FALSE]
cat("Genes kept after filtering:", nrow(dge), "\n")

dge <- calcNormFactors(dge, method = "TMM")
logcpm <- cpm(dge, log = TRUE, prior.count = 2)

lib_tab <- data.frame(
  sample = colnames(dge),
  lib_size = dge$samples$lib.size,
  norm_factor = dge$samples$norm.factors,
  condition = pheno_clean$condition
)

knitr::kable(head(lib_tab, 10),
             caption = "Table 9.1: Library sizes and TMM normalization factors (first 10 samples).")
```

## Pre vs post normalization plots

```{r pre_post_plots}
par(mfrow = c(1,2))

boxplot(log2(dge$counts + 1),
        main = "Figure 9.1: Pre-normalization\nlog2(counts+1)",
        outline = FALSE,
        las = 2)

boxplot(logcpm,
        main = "Figure 9.2: Post-normalization\nlogCPM (TMM)",
        outline = FALSE,
        las = 2)

par(mfrow = c(1,1))
```

**Figure 9.1–9.2 legend:**  
Boxplots summarize sample-wise expression distributions before and after normalization. After TMM normalization (logCPM), distributions are more comparable across samples, supporting reduced technical variability.

```{r norm_density}
plot(density(logcpm[,1]),
     main="Figure 9.3: Density of logCPM after TMM",
     xlab="logCPM")

for(i in 2:ncol(logcpm)){
  lines(density(logcpm[,i]), col=i)
}
```

**Figure 9.3 legend:**  
Density curves of normalized logCPM values. Improved alignment across samples indicates that normalization has reduced library-driven differences while preserving variation needed for biological comparisons.

# Structure and outlier assessment (MDS + PCA)

Exploratory ordination methods provide a global view of whether samples cluster by condition and whether any samples appear inconsistent with their labels.

```{r mds_plot}
plotMDS(dge,
        labels = pheno_clean$condition,
        col = as.integer(pheno_clean$condition))
title("Figure 10.1: MDS plot (edgeR)")

legend("topright",
       legend = levels(pheno_clean$condition),
       col = 1:length(levels(pheno_clean$condition)),
       pch = 16,
       bty = "n")
```

**Figure 10.1 legend:**  
MDS summarizes sample-to-sample distances based on expression profiles. Separation between IPF and Control suggests condition-driven structure. Samples that fall far from their group would be investigated as potential outliers.

```{r pca_plot}
pca <- prcomp(t(logcpm), scale. = TRUE)

plot(pca$x[,1], pca$x[,2],
     col = as.integer(pheno_clean$condition),
     pch = 19,
     xlab = "PC1",
     ylab = "PC2",
     main = "Figure 10.2: PCA of normalized expression")

legend("topright",
       legend = levels(pheno_clean$condition),
       col = 1:length(levels(pheno_clean$condition)),
       pch = 19,
       bty = "n")
```

**Figure 10.2 legend:**  
PCA displays major axes of variability in the normalized dataset. If IPF vs Control separates along early PCs, this supports that disease status explains a meaningful fraction of expression variance.

**Outliers removed (explicit answer):**  
No samples were removed automatically. Outlier removal is only justified if there is evidence of technical error or mislabeling; otherwise biological heterogeneity is expected in human patient cohorts.

# Differential expression analysis (limma-voom)

Because RNA-seq count data exhibit a mean–variance relationship, limma-voom is used to estimate precision weights and enable linear modeling with empirical Bayes moderation. This approach is well-established for RNA-seq differential expression and performs reliably with moderate sample sizes.

```{r de_analysis}
design <- model.matrix(~ 0 + pheno_clean$condition)
colnames(design) <- levels(pheno_clean$condition)

v <- voom(dge, design, plot = TRUE)

fit <- lmFit(v, design)
cont <- makeContrasts(IPFvsControl = IPF - Control, levels = design)
fit2 <- eBayes(contrasts.fit(fit, cont))

tt <- topTable(fit2, number = Inf, sort.by = "P")
tt$gene <- rownames(tt)

knitr::kable(head(tt, 10),
             caption = "Table 11.1: Top 10 differentially expressed genes (IPF vs Control).")
```

**Voom plot legend:**  
The voom mean–variance plot visualizes the dependence of variance on mean expression and the fitted trend used to compute observation-level weights. This step enables appropriate linear modeling of RNA-seq derived expression values.

# Significance thresholds and multiple testing (explicit)

To define a practical and interpretable set of differential expression hits, I used two criteria: statistical significance after multiple testing correction and a minimum effect size. Specifically:

- **FDR < 0.05** using the Benjamini–Hochberg procedure  
- **|log2 fold-change| ≥ 1** to focus on changes of at least ~2-fold  

FDR control is necessary because thousands of genes are tested simultaneously; without correction, many low p-values would occur by chance (Benjamini & Hochberg, 1995).

```{r thresholds}
tt$signif <- tt$adj.P.Val < 0.05 & abs(tt$logFC) >= 1
cat("DE genes (FDR<0.05 & |logFC|>=1):", sum(tt$signif), "\n")
```

# Visualizations (Volcano, MA, Heatmap)

```{r volcano}
ggplot(tt, aes(x = logFC, y = -log10(adj.P.Val))) +
  geom_point(aes(color = signif), alpha = 0.7) +
  theme_minimal() +
  labs(
    title = "Figure 13.1: Volcano plot (IPF vs Control)",
    x = "log2 Fold Change (IPF - Control)",
    y = "-log10(FDR)"
  )
```

**Figure 13.1 legend:**  
Volcano plot combines effect size (log2FC) and statistical significance (FDR). Highlighted points indicate genes passing both FDR and effect size thresholds.

```{r ma_plot}
plotMA(fit2, main = "Figure 13.2: MA plot (IPF vs Control)")
```

**Figure 13.2 legend:**  
MA plot shows log fold-change versus average expression. It helps assess whether differential expression is concentrated in low-expression genes and provides a complementary view to the volcano plot.

```{r heatmap}
# pick top genes safely
top50 <- tt$gene[order(tt$adj.P.Val)][1:50]
mat_top <- logcpm[top50, , drop = FALSE]

# Build annotation_col with rownames matching the matrix columns
annotation_col <- data.frame(condition = pheno_clean$condition)
rownames(annotation_col) <- colnames(mat_top)

# Ensure condition is a factor with correct levels
annotation_col$condition <- factor(annotation_col$condition, levels = c("Control", "IPF"))

# Provide explicit colors for annotation to avoid 'fill length 0' errors
annotation_colors <- list(
  condition = c(Control = "#4DAF4A", IPF = "#E41A1C")
)

pheatmap::pheatmap(
  mat_top,
  scale = "row",
  annotation_col = annotation_col,
  annotation_colors = annotation_colors,
  show_colnames = FALSE,
  main = "Figure 13.3: Top 50 DE genes (row-scaled logCPM)"
)
```
**Figure 13.3 legend:**  
Heatmap of the 50 most significant genes (by FDR). Row-scaling emphasizes relative expression patterns across samples. As observed in the plot the samples cluster distinctly by condition (IPF vs Control), indicating a coherent disease-associated signature. The IPF samples with red annotation show consistent upregulation of a specific gene subset compared to controls, validating biologilical distinctness of the test group.

# Short biological interpretation (course-appropriate)

The differential expression results suggest systematic transcriptional differences between IPF and control macrophage-rich BAL samples. While this assignment focuses on producing a clean and normalized dataset, the observed structure in MDS/PCA and the DE signal are consistent with the expectation that IPF involves altered immune activation and tissue remodeling programs. In subsequent assignments, gene set enrichment and pathway scoring will provide higher-level biological interpretation (e.g., whether extracellular matrix organization, interferon signaling, or profibrotic pathways are enriched).

# Final dataset object (required output specification)

The required deliverable is a normalized expression matrix with unique HUGO symbols as rownames and numeric expression values across samples as columns. Here I store the final dataset as `final_dataset` and verify the required properties.

```{r final_dataset}
final_dataset <- logcpm

stopifnot(length(unique(rownames(final_dataset))) == nrow(final_dataset))
stopifnot(is.numeric(final_dataset[1,1]))

cat("Final dataset dims (genes x samples):", paste(dim(final_dataset), collapse = " x "), "\n")

knitr::kable(
  head(final_dataset[, 1:5]),
  caption = "Table 15.1: Preview of final normalized dataset (logCPM; TMM-normalized)."
)
```

# Explicit answers to required interpretation questions

**Why is the dataset of interest to you?**  
This dataset is of interest because it provides human macrophage-rich BAL transcriptomes in a clinically meaningful contrast (IPF vs smoking control). IPF is an immune-influenced fibrotic disease, and macrophages are central regulators of inflammation and remodeling. As a result, macrophage transcriptomic differences between disease and control conditions are a plausible and informative readout of fibrosis-associated immune programs.

**What are the control and test conditions?**  
The control condition is smoking control BAL samples, and the test condition is IPF BAL samples.

**How many samples are in each condition?**  
The sample counts are reported in Table 6.1 (Control vs IPF) as parsed from GEO metadata.

**Were there expression values that were not unique for specific genes? How did you handle these?**  
Yes. Multiple rows can map to the same gene symbol after HGNC cleanup. I handled this by collapsing duplicate gene symbols using `rowsum()` (summing counts), yielding one row per gene and producing a unique-symbol expression matrix compatible with downstream analysis.

**Were there expression values that could not be mapped to current HUGO symbols?**  
Yes. Rows with empty or missing gene symbols after HGNChelper cleanup were removed to avoid ambiguous identifiers.

**Were there any outliers? How were they handled? How many outliers were removed?**  
Outliers were assessed using global structure plots (MDS and PCA). In my analysis, no samples were removed as they showed reasonable clustering by group. Regarding the originating paper (Prasse et al., 2018), the authors analyzed the same 23 samples (16 IPF, 7 Control) present in this GEO dataset. They did not report excluding any samples from this final set for quality reasons.

**How did you handle replicates?**  
Replicates are biological samples within each condition. They are handled through the group factor in the design matrix and contribute to variance estimation in the differential expression model.

**What is the final coverage of your dataset?**  
Final coverage is the number of genes after mapping, duplicate collapsing, filtering, and normalization: `r nrow(final_dataset)` genes.

# Conclusion

This notebook implements a complete, reproducible workflow to prepare a clean, normalized human RNA-seq expression dataset for subsequent BCB420 assignments. The dataset was downloaded computationally from GEO, assessed for basic quality characteristics, mapped to current HUGO gene symbols, cleaned through removal of unmappable identifiers and collapsing of duplicates, filtered to remove lowly expressed genes, and normalized using TMM. Exploratory plots demonstrate that normalization improves comparability across samples while preserving condition-associated structure. Differential expression analysis using limma-voom identifies genes associated with IPF status in macrophage-rich BAL samples. The final deliverable is a normalized expression matrix with unique gene symbols and numeric expression values suitable for downstream scoring and pathway analysis.

# References

1.  **Dataset (Originating Paper):** Prasse A, Binder H, Schupp JC, Kayser G, Bargagli E, Jaeger B, et al. (2018). BAL Cell Gene Expression in Idiopathic Pulmonary Fibrosis: A Pilot Study. *PLOS ONE*, 14(1): e0210463.

2.  **limma:** Ritchie ME, Phipson B, Wu D, Hu Y, Law CW, Shi W, Smyth GK. (2015). limma powers differential expression analyses for RNA-sequencing and microarray studies. *Nucleic Acids Research*, 43(7):e47.

3.  **edgeR (TMM Normalization):** Robinson MD, McCarthy DJ, Smyth GK. (2010). edgeR: a Bioconductor package for differential expression analysis of digital gene expression data. *Bioinformatics*, 26(1):139–140.

4.  **GEOquery:** Davis S, Meltzer PS. (2007). GEOquery: a bridge between the Gene Expression Omnibus (GEO) and BioConductor. *Bioinformatics*, 23(14):1846–1847.

5.  **HGNChelper:** Xiao W, Adjeroh D, Powell JI. (2010). HGNChelper: identification and correction of gene symbols. *BMC Bioinformatics*, 11:539.

6.  **pheatmap:** Kolde R. (2019). pheatmap: Pretty Heatmaps. R package version 1.0.12. https://CRAN.R-project.org/package=pheatmap

7.  **FDR Method:** Benjamini Y, Hochberg Y. (1995). Controlling the false discovery rate: a practical and powerful approach to multiple testing. *Journal of the Royal Statistical Society: Series B*, 57(1):289–300.
